Bioconductor workflow for single-cell RNA sequencing: Normalization, dimensionality reduction, clustering, and lineage inference

### 2017-07-06 workflow/workflow.Rmd
temptatives to change colors in plotbarplot

- make sure versions package ok and pushed to bioc (DR, EP, FP),
- update clustering section with new clusterExperiment functions (EP),
- address Sandrine's comments in sections slingshot and de (KS),
- try running workflow with more genes (DR),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package using cache instead of saved files for zinbwave and clusterExperiment and making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.pdf")
```

@@

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering."}
plotBarplot(ceObj, legend = F) 
```

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figures \@ref(fig:addPublishedClusters) and \@ref(fig:comparePublished)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for original published clustering."}
ceObj<-addClusters(ceObj, colData(ceObj)$publishedClusters, 
                   clusterLabel = "publishedClusters")
plotBarplot(ceObj,whichClusters=c("combineMany","publishedClusters"), legend = F) 
```

```{r comparePublished, fig.cap="RSEC: Comparison between the clustering found using our workflow and the original published clustering."}
plotClusters(ceObj, whichClusters=c("mergeClusters","combineMany","publishedClusters"))
```

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | mOSN | orange | original 9, 12
c5 | Immature Neuron | purple | original 10, 14
c6 | MV | brown | original 15

@@

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl = our_cl[our_cl != "-1"]
pal = pal[names(pal) != "-1"]
X <- W[our_cl != "-1", ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)

pairs(lineages, type="curves", col = pal[cl])
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c8` and `c6` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c6"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assays(se)$normalizedValues[, our_cl != "-1"]
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```

>>>>

- try running workflow with more genes (DR),
- clean up singlecellworkflow package (FP),
- update NAMESPACE and DESCRIPTION (FP),
- install and run package making sure we can build the vignette using R 3.4 and bioc-devel 3.6 (all?),
- make singlecellworkflow package public (FP).]

@@

```{r zinbschema, echo=FALSE, out.width="95%", fig.cap="ZINB-WaVE: Schematic view of the ZINB-WaVE model. This figure was reproduced with kind permission from Risso et al. (2017)."}
knitr::include_graphics("zinb_schema.png")
```

@@

```{r barplotOurs, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by RSEC clustering."}
plotBarplot(ceObj, legend = F) 
```

The distribution of cells in our workflow's clustering overall agrees with that in the original published clustering (Figure \@ref(fig:addPublishedClusters)), the main difference being that several of the published clusters were merged here into single clusters. This discrepancy is likely caused by the fact that we performed dimensionality reduction using ZINB-WaVE and provided the resulting `W` matrix as input to `RSEC` instead of performing dimensionality reduction on the normalized counts within the `RSEC` function.

[FP: Elizabeth or Davide, how can we change the colors in Figure \@ref(fig:addPublishedClusters) to match the colors we use for publishedClusters (that is `col_clus` like in Figure \@ref(fig:mdsW)). I tried colPalette argument, but it does not work with SummarizedExperiment input. Also, would it be possible to have the same order and labels for the barplots in Figures \@ref(fig:barplotOurs) and \@ref(fig:addPublishedClusters) (barplots labelled -1 and Not Assigned are the same, right?)?

```{r addPublishedClusters, fig.cap="RSEC: Barplot of number of cells per cluster, for our workflow's clustering, color-coded by original published clustering."}
ceObj <- addClusters(ceObj, colData(ceObj)$publishedClusters, 
                     clusterLabel = "publishedClusters")
plotBarplot(ceObj, whichClusters=c("combineMany","publishedClusters"),
            xlab = "", legend = F) 
```

@@

Cluster name | Description | Color | Correspondence
-------------|-------------|-------| ----------
c1 | HBC | blue | original 1, 5
c2 | GBC | green | original 2, 3, 11
c3 | mSUS | red | original 4, 7
c4 | Contaminants | orange | original -2 
c5 | mOSN | purple | original 9, 12
c6 | Immature Neuron | brown | original 10, 14
c7 | MV | cian | original 15

Cells in cluster `c4` have a cluster label of `-2` in the original published clustering meaning that they were not assigned to any cluster in the original publication. These cells were actually identified as non-sensory contaminants as they overexpress gene `Reg3g` (see Figure S1 from [@Fletcher2017] and Figure \@ref(fig:boxplotReg3g)) and were removed from the original published clustering. While it is reassuring that our workflow clustered these cells separately with no influence on the clustering of the other cells, we removed cluster `c4` to infer lineages and pseudotimes as cells in this cluster do not participate in the cell differentiation process. Note that our workflow's clustering did not capture all the cells overexpressing gene `Reg3g` in cluster `c4` (see Figure \@ref(fig:boxplotReg3g)), however we keep these cells to infer lineages as they did not seem to influence the clustering.

[FP: check writing + table with correspondance of clusters]

```{r boxplotReg3g, fig.width=4, fig.height=4, fig.cap="Boxplots of the log counts of gene Reg3g for cells in cluster c4 (left boxplot) and cells not in cluster c4 (right boxplot)."}
c4 <- rep("other clusters", ncol(se)) 
c4[primaryClusterNamed(ceObj) == "c4"] <- "cluster c4"
boxplot(log1p(assay(se)["Reg3g", ]) ~ c4, ylab = "log counts", cex.axis = .8, cex.lab = .8)
```

@@

```{r}
our_cl <- primaryClusterNamed(ceObj)
cl <- our_cl[!our_cl %in% c("-1", "c4")]
pal <- pal[!names(pal) %in% c("-1", "c4")]
X <- W[!our_cl %in% c("-1", "c4"), ]
mds <- cmdscale(dist(X), eig = TRUE, k = 4)
X <- mds$points

@@

```{r curves, fig.cap="Slingshot: Cells color-coded by cluster in a 4-dimensional MDS space with smooth curves representing each inferred lineage."}
lineages <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[cl])
```

@@

In the workflow, we recover a reasonable ordering of the clusters using the unsupervised version of slingshot. However, in some other cases, we have noticed that we need to give more guidance to the algorithm to find the correct ordering. `getLineages` has the option for the user to provide known end cluster(s). Here is the code to use `slingshot` in a supervised setting, where we know that clusters `c3` and `c7` represent terminal cell fates.

```{r slingshotsupervised, eval=FALSE}
lineages <- getLineages(X, clusterLabels = cl, start.clus = "c1",
                         end.clus = c("c3", "c7"))
lineagees <- getCurves(lineages)
pairs(lineages, type="curves", col = pal[primaryClusterNamed(ceObj)])
pairs(lineages, type="lineages", col = pal[primaryClusterNamed(ceObj)], show.constraints = TRUE)

@@

```{r fitgam}
t <- pseudotime(lineages)[,1]
y <- assays(se)$normalizedValues[, !our_cl %in% c("-1", "c4")]
gam.pval <- apply(y,1,function(z){
  d <- data.frame(z=z, t=t)
  tmp <- gam(z ~ lo(t), data=d)
  p <- summary(tmp)[4][[1]][1,5]
  p
})
```
